Trait plasticity and covariance along a continuous soil moisture gradient

### 2021-01-17 FVT Manuscript/manuscript.tex
V5

\documentclass[jou,floatsintext]{apa6}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage{hyperref}
\hypersetup{unicode=true,
            pdftitle={Trait plasticity and covariance along a continuous soil moisture gradient},
            pdfauthor={J. Grey Monroe, Haoran Cai, \& David L. Des Marais},
            pdfkeywords={drought, function-valued traits, Brachypodium, evolutionary constraint, phenotypic plasticity, non-linearity},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

%%% Use protect on footnotes to avoid problems with footnotes in titles
\let\rmarkdownfootnote\footnote%
\def\footnote{\protect\rmarkdownfootnote}

  \title{Trait plasticity and covariance along a continuous soil moisture gradient}
    \author{J. Grey Monroe\textsuperscript{1}, Haoran Cai\textsuperscript{2}, \& David L. Des Marais\textsuperscript{2,3}}
    \date{}

\shorttitle{Function-valued drought extremes}
\affiliation{
\vspace{0.5cm}
\textsuperscript{1} Department of Plant Sciences, University of California at Davis, Davis, USA\\\textsuperscript{2} Department of Civil and Environmental Engineering, Massachusetts Institute of Technology, Cambridge, MA, USA\\\textsuperscript{3} The Arnold Arboretum of Harvard University. Boston, MA, USA}
\keywords{drought, function-valued traits, Brachypodium, evolutionary constraint, phenotypic plasticity, non-linearity}
\usepackage{csquotes}
\usepackage{upgreek}
\captionsetup{font=singlespacing,justification=justified}

@@

@@

Correspondence concerning this article should be addressed to J. Grey Monroe, 262 Robbins Hall, UC Davis, One Shields Ave, Davis CA 95616. E-mail: \href{mailto:greymonroe@gmail.com}{\nolinkurl{greymonroe@gmail.com}}}

\abstract{
Water is perhaps the greatest environmental determinant of plant yield and fitness, yet our understanding of plant-water relations is limited because it is primarily informed by experiments that have treated soil moisture variability as two simple discrete levels - wet and dry - rather than a continually varying environmental gradient. Here we used experimental and statistical methods based on function-valued traits to study responses to continuously varying soil moisture gradient in a suite of physiological and morphological traits in two species and multiple genotypes of the model grass \textit{Brachypodium}. We find that the majority of traits exhibit non-linear responses to soil moisture variability. We also observe differences in the shape of these non-linear responses between traits, species, and genotypes. Emergent phenomena arise from this variation including changes in trait correlations and evolutionary constraints as a function of soil moisture. These results point to the importance of considering non-linearity in plant-water relations to predict plastic and evolutionary responses to changing climates.

@@

For plants, soil water availability is one of the most important environmental factors in ecology and agriculture, acting as a major determinant of fitness and yield (Juenger 2013; Greenham \emph{et al.} 2017). Considerable interest and effort have been placed on studying plant responses to drought, often defined conceptually and experimentally as an environmental condition of abnormally elevated aridity resulting in decreased plant performance (Passioura 1996). Most of the vast research on drought responses and tolerance strategies, including in \emph{Brachypodium}, the focal system of this work, has been limited to comparisons between discrete soil water levels - control and water-limited (Des Marais \emph{et al.} 2012; Edwards \emph{et al.} 2012; El-Soda \emph{et al.} 2014; Vasseur \emph{et al.} 2014; Greenham \emph{et al.} 2017). Yet soil moisture as an environmental factor is complex and multidimensional, with fluctuations varying continuously in timing, duration, and degree. Here we investigate trait responses to one important dimension of soil moisture variability - degree - with experimental and statistical approaches treating soil moisture content as a continuous variable rather than a set of fixed levels.

The limited extent to which continuous environmental variables have been studied is not unique to investigations of plant responses to soil moisture. Recent decades have seen increased appreciation for modeling traits as mathematical functions of continuous variables across diverse organisms (Pettay \emph{et al.} 2008; Rocha and Klaczko 2012; Mason \emph{et al.} 2020), a wide range of traits (Robinson \emph{et al.} 2009; McGuigan \emph{et al.} 2010; Stinchcombe \emph{et al.} 2010), and numerous components of the environment (Brommer \emph{et al.} 2008; McGuigan 2009; Pearse \emph{et al.} 2019). This burgeoning interest in the importance of \enquote{function-valued traits} has evolved from a call-to-arms to a field experiencing rapid theoretical and empirical advances (Kirkpatrick and Heckman 1989; Kingsolver \emph{et al.} 2001; Griswold \emph{et al.} 2008; Stinchcombe \emph{et al.} 2012; Goolsby 2015; Gomulkiewicz \emph{et al.} 2018). For plant biologists this presents a compelling framework for gaining a deeper understanding of plant adaptations to the environment.

A deep body of literature aims to understand the traits and strategies used by plants to acclimate to variable soil water content. Quantitative genetic variation in myriad drought resistance traits have been observed in natural populations and laboratory model systems. In particular, natural populations of Brassicaceae species (including \emph{Arabidopsis thaliana}) and \emph{Brachypodium} harbor variation both constitutive and inducible traits mediating plant-water relations, including water use efficiency (Des Marais \emph{et al.} 2012, 2017; Edwards \emph{et al.} 2012; Greenham \emph{et al.} 2017), leaf chemistry (Kesari \emph{et al.} 2012; Des Marais \emph{et al.} 2017), leaf anatomy (Skirycz \emph{et al.} 2011; Verelst \emph{et al.} 2013; Dittberner \emph{et al.} 2018), root-shoot biomass partitioning (Des Marais \emph{et al.} 2012, 2017), and many others (Verslues and Juenger 2011; Edwards \emph{et al.} 2012; Juenger 2013; Luo \emph{et al.} 2016; Yarkhunova \emph{et al.} 2016; Lenk \emph{et al.} 2019). Among the numerous plant traits that can and have been measured, few have been studied more extensively than specific leaf area (SLA - often reported as it's inverse, leaf mass per area or LMA). SLA provides a description of leaf architecture that is central to the leaf economics spectrum, a theory which seeks to explain variation in leaf physiological strategies, from more conservative (low SLA) to more productive (high SLA) (Wright \emph{et al.} 2004). In the context of drought stress, it is intuitive that low SLA might be adaptive as lower leaf surface area is expected to reduce water loss through transpiration. In several cases, reductions in SLA have been reported under drought conditions (Casper \emph{et al.} 2001). For all of these traits, the shape of plastic responses to variation in water availability remain largely unmeasured.

Despite abundant evidence of natural variation in organism-environment interaction, little is known as to how trait covariance changes across continuous environmental gradients (but see Robinson \emph{et al.} 2009). Describing this structure is important as trait variances and covariances can result in differential evolutionary constraints in alternative environments. For example, if total genetic variance in trait space changes depending on the environment, then the capacity to respond to selection will vary accordingly, with reduced responses to selection under conditions where trait variation is lower and vice versa. Similarly, if trait covariances depend on environment, conditions which increase trait covariances may limit evolutionary potential across a range of environments by reducing the effective axes of variation (Levins 1968; Via and Lande 1985; Kingsolver and Gomulkiewicz 2003; Gomulkiewicz \emph{et al.} 2018). These phenomena are made more complex by the possibility that the relationship between trait variances and covariances with the environment may be non-linear. Investigating complexities of genetic architecture is therefore needed for developing more complete quantitative predictions of evolvability in rapidly changing environments. Fortunately, advances have been made in statistical methods to reduce the dimensionality of genetic variance covariance matrices and produce meaningful summaries describing evolutionary constraints (Houle 1992; Blows 2007; Kirkpatrick 2009; Kingsolver \emph{et al.} 2015).

In this study we combined these approaches with function-valued trait analyses to investigate trait plasticity, covariance and evolutionary constraints across soil moisture gradients. We model trait responses to a continuous gradient of soil moisture for a range of genotypes of two species of the model grass genus \emph{Brachypodium}. Because the shapes of trait responses cannot necessarily be known \emph{a priori}, we use model selection among linear and non-linear environmental predictors to estimate the response function for each trait. We then estimate genotype means at different levels of soil moisture and compute the variance-covariance parameters for all traits. Finally, we ask whether patterns of variance and covariance in drying-responsive traits in \emph{Brachypodium} species may lead to variation in evolutionary constraint as a function of soil moisture.

@@

\emph{Brachypodium} is a model genus for the genetics and genomics of C3 grasses (Brkljacic \emph{et al.} 2011). In this work we studied natural variation between and among two species of \emph{Brachypodium}: the annual \emph{B. distachyon} and the perennial \emph{B. sylvaticum}. Both species are endemic to Eurasia, with \emph{B.distachyon} more prevalent in seasonally dry habitats in Southern Europe, North Africa, and the Middle East, and \emph{B. sylvaticum} more widely distributed throughout Eurasia (Catalan \emph{et al.} 2016) (Figure \ref{fig:maps}).

Five genotypes of each species were studied to characterize patterns of variation in plant traits across an environmental gradient: \emph{Brachypodium distachyon} inbred lines ABR2, Adi-10, Bd21, Bd3-1, and Koz-1 and \emph{Brachypodium sylvaticum} inbred lines Ain-1, Ast-1, Kry-1, Osl-1, Vel-1. For each species, these genotypes represent a range of geographical origins and phenotypic diversity (Steinwand \emph{et al.} 2013; Des Marais \emph{et al.} 2017). Both species are self-compatible and each of the lines used here have been maintained as inbred lines for greater than six generations (Vogel \emph{et al.} 2009; Steinwand \emph{et al.} 2013); as such, experimental replicates may be considered nearly homozygous.

@@

Dry down treatments began 29 and 42 days after sowing (DAS) for \emph{B. distachyon} and \emph{B. sylvaticum}, respectively. Because the harvesting was divided over five consecutive days (see section below), plants were split into five equal harvest cohorts, with each cohort containing equal numbers of each watering treatment to avoid confounding harvest day with soil moisture content. Thus, though each consecutive cohort differed in age by a single day, each experienced the dry down treatment for the same amount of time. Nevertheless, we expected the difference in age between harvest cohorts to potentially impact trait expression and we therefore included harvest day (cohort) as a covariate in subsequent models. To generate a continuous gradient of final soil moisture by the end of the dry down period, plants were split into five watering treatments, receiving approximately 0, 4, 8, 12, 16, or 20 ml of water per day for 14 days. Prior to initiating the experiment pots were weighed with dry soil (\(mass_{dry}\)) and field capacity (\(mass_{max}\)) at the beginning of the experiment. These measurements provide the basis for calculating the gravimetric soil moisture content on each day and at the conclusion of the experiment. During the course of the dry down experiment, soil moisture content was calculated during the morning of day \(d\) for each pot as \(mass_d/(mass_{max}-mass_{dry})\).

@@

To characterize phenotypic responses to our experimental soil moisture gradient, we measured a suite of developmental and physiological traits. Plants were harvested in five cohorts over five days. Each day, half of the sampled plants were harvested for above and below ground biomass, total above ground green area, \(\delta C_{13}\), N content, C content. The other half were harvested and assessed for specific leaf area (SLA) and relative water content (RWC). Above ground leaf area was estimated by laying freshly harvested plants flat between plates of clear plexiglass and imaging with a Nikon 5300 digital camera at a fixed distance with a 35mm Nikkor lens. Total green pixels were counted for each image with Easy Leaf Area (\url{https://github.com/heaslon/Easy-Leaf-Area}) with settings shown in Figure \ref{fig:leafarea}. Above ground biomass was measured after drying leaf material overnight at 60°C and then for several weeks at room temperature. Below ground biomass was measured after washing the soil matrix from roots and drying them overnight at 60°C and then for several weeks at room temperature. Above and belowground biomass was measured after leaves and roots were dried. Leaf tissues for \(\delta C_{13}\), \(\delta N_{15}\), nitrogen (hereafter \enquote{N}) content, and carbon (hereafter \enquote{C}) content were ground to a fine powder and processed by the UC Davis Stable Isotope Facility. SLA was calculated by scanning the two youngest fully emerged leaves with a 1 \(cm^2\) red square. Leaf area in \(mm^2\) was calculated from these same images using Easy Leaf Area. SLA was calculated as \(leaf\ area / biomass_{dry}\). These leaves were also used to calculate RWC. Prior to drying, fresh leaves were weighed (\(biomass_{fresh}\)) and then submerged under water in 15mL falcon tubes for several hours. They were then weighed (\(biomass_{turgid})\), oven-dried overnight, and weighed again (\(biomass_{dry}\)). RWC was calculated as \((biomass_{fresh}-biomass_{dry})/(biomass_{turgid}-biomass_{dry})\).

\begin{table*}[tbp]
\begin{center}
\begin{threeparttable}
\caption{\label{tab:models}Model selection. *** = predictor variable p < 1e-5, ** = predictor variable p < 1e-3, * = predictor variable p < 0.5, - = predictor variable included in selected model but p > 0.05. H = harvest day, S = species, G = genotype, E = final soil moisture, E\textasciicircum{}2 = quadratic parameter, ns(E) = 2nd degree natural spline parameter}
\scriptsize{
\begin{tabular}{ccccccccc}
\toprule
 & \multicolumn{1}{c}{G} & \multicolumn{1}{c}{E} & \multicolumn{1}{c}{E\textasciicircum{}2} & \multicolumn{1}{c}{ns(E)} & \multicolumn{1}{c}{H} & \multicolumn{1}{c}{G*E} & \multicolumn{1}{c}{G*E\textasciicircum{}2} & \multicolumn{1}{c}{G*ns(E)}\\
\midrule
RWC - B. distachyon &  &  & *** & *** & * &  &  & \\ \midrule
RWC - B. sylvaticum & * &  &  & *** & * &  &  & \\ \midrule
SLA - B. distachyon & - &  & *** & *** & *** &  & * & \\ \midrule
SLA - B. sylvaticum & *** &  &  & - &  &  &  & \\ \midrule
Green Area - B. distachyon & ** &  &  & *** & *** &  &  & \\ \midrule
Green Area - B. sylvaticum & *** &  & * &  & ** &  & - & \\ \midrule
Shoot Mass - B. distachyon & *** &  & *** &  & *** &  &  & \\ \midrule
Shoot Mass - B. sylvaticum & *** &  & - &  & * &  &  & \\ \midrule
Root Mass - B. distachyon & *** &  & *** & - & * &  &  & -\\ \midrule
Root Mass - B. sylvaticum & *** &  &  &  &  &  &  & \\ \midrule
Root:Shoot - B. distachyon & *** &  &  & * & *** &  &  & \\ \midrule
Root:Shoot - B. sylvaticum & *** &  & * &  & *** &  & * & \\ \midrule
Biomass - B. distachyon & *** &  &  & *** & *** &  &  & \\ \midrule
Biomass - B. sylvaticum & *** &  &  &  & * &  &  & \\ \midrule
C content - B. distachyon &  &  & * &  &  &  &  & \\ \midrule
C content - B. sylvaticum & * &  &  & - &  &  &  & \\ \midrule
d13c - B. distachyon & *** &  & *** &  & *** &  &  & \\ \midrule
d13c - B. sylvaticum & *** &  &  & ** & *** &  &  & \\ \midrule
N content - B. distachyon & *** &  & - &  & * &  &  & \\ \midrule
N content - B. sylvaticum & *** &  &  & *** &  &  &  & **\\ \midrule
d15n - B. distachyon & ** &  &  & ** & *** &  &  & \\ \midrule
d15n - B. sylvaticum &  &  & *** & - & ** &  &  & \\ \midrule
C:N ratio - B. distachyon & *** &  & ** &  & * &  &  & \\ \midrule
C:N ratio - B. sylvaticum & *** &  & *** &  &  &  & * & \\ \midrule
\bottomrule
\end{tabular}
}
\end{threeparttable}
\end{center}
\end{table*}

@@

\[v_T = \sum_{n=1}^{n} \lambda_i\]

>>>>

\PassOptionsToPackage{unicode=true}{hyperref} % options for packages loaded elsewhere
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[jou,floatsintext]{apa6}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provides euro and other symbols
\else % if luatex or xelatex
  \usepackage{unicode-math}
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage[]{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\usepackage{hyperref}
\hypersetup{
            pdftitle={Trait plasticity and covariance along a continuous soil moisture gradient},
            pdfauthor={J. Grey Monroe, Haoran Cai, \& David L. Des Marais},
            pdfkeywords={drought, function-valued traits, Brachypodium, evolutionary constraint, phenotypic plasticity, non-linearity},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

% set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother

\shorttitle{Function-valued drought extremes}
\affiliation{
\vspace{0.5cm}
\textsuperscript{1} Department of Plant Sciences, Univeƒrsity of California at Davis, Davis, USA\\\textsuperscript{2} Department of Civil and Environmental Engineering, Massachusetts Institute of Technology, Cambridge, MA, USA\\\textsuperscript{3} The Arnold Arboretum of Harvard University. Boston, MA, USA\\\textsuperscript{4} Max Planck Institute for Developmental Biology, Tubingen, DE}
\keywords{drought, function-valued traits, Brachypodium, evolutionary constraint, phenotypic plasticity, non-linearity}
\usepackage{csquotes}
\usepackage{upgreek}
\captionsetup{font=singlespacing,justification=justified}

@@

\title{Trait plasticity and covariance along a continuous soil moisture gradient}
\author{J. Grey Monroe\textsuperscript{1,4}, Haoran Cai\textsuperscript{2}, \& David L. Des Marais\textsuperscript{2,3}}
\date{}

@@

Correspondence concerning this article should be addressed to J. Grey Monroe, 150 California Ave, Davis, CA 95616. E-mail: \href{mailto:gmonroe@ucdavis.edu}{\nolinkurl{gmonroe@ucdavis.edu}}}

\abstract{
Water availality is perhaps the greatest environmental determinant of plant yield and fitness. Yet our understanding of plant-water relations is limited because it is primarily informed by experiments considering soil moisture variability at two discrete levels - wet and dry - rather than as a continually varying environmental gradient. Here we used experimental and statistical methods based on function-valued traits to explore responses to continuously varying soil moisture gradient in physiological and morphological traits in two species and five genotypes each of the model grass \textit{Brachypodium}. We find that most traits exhibit non-linear responses to soil moisture variability. We also observe differences in the shape of these non-linear responses between traits, species, and genotypes. Emergent phenomena arise from this variation including changes in trait correlations and evolutionary constraints as a function of soil moisture. These results point to the importance of considering non-linearity in plant-water relations to understand plastic and evolutionary responses to changing climates.

@@

For plants, soil water availability is an important environmental factor in ecology and agriculture, acting as a major determinant of fitness and yield (Juenger 2013; Greenham \emph{et al.} 2017). As such, considerable effort has been placed on studying plant responses to drought, often defined conceptually and experimentally as an environmental condition of abnormally elevated aridity resulting in decreased plant performance (Passioura 1996). Most of the research on drought responses and tolerance strategies, including in \emph{Brachypodium}, the focal system of this work, has involved comparisons between discrete soil water levels - control and water-limited (Des Marais \emph{et al.} 2012; Edwards \emph{et al.} 2012; El-Soda \emph{et al.} 2014; Vasseur \emph{et al.} 2014; Greenham \emph{et al.} 2017). Yet soil moisture as an environmental factor is complex and multidimensional, with fluctuations varying continuously in timing, duration, and degree. Here we investigate trait responses to one important dimension of soil moisture variability - degree - with experimental and statistical approaches treating soil moisture content as a continuous variable rather than a set of fixed levels.

Quantitative genetic variation in drought resistance traits have been observed in natural populations and laboratory model systems. In particular, natural populations of Brassicaceae species (including \emph{Arabidopsis thaliana}) and \emph{Brachypodium} harbor variation both constitutive and plastic traits mediating plant-water relations, including water use efficiency (Des Marais \emph{et al.} 2012, 2017; Edwards \emph{et al.} 2012; Greenham \emph{et al.} 2017), leaf chemistry (Kesari \emph{et al.} 2012; Des Marais \emph{et al.} 2017), leaf anatomy (Skirycz \emph{et al.} 2011; Verelst \emph{et al.} 2013; Dittberner \emph{et al.} 2018), root-shoot biomass partitioning (Des Marais \emph{et al.} 2012, 2017), and many others (Verslues and Juenger 2011; Edwards \emph{et al.} 2012; Juenger 2013; Luo \emph{et al.} 2016; Yarkhunova \emph{et al.} 2016; Lenk \emph{et al.} 2019). Among the many plant traits that can and have been measured, few have been studied more extensively than specific leaf area (SLA - often reported as its inverse, leaf mass per area or LMA). SLA provides a description of leaf architecture that is central to the leaf economics spectrum, a theory which seeks to explain variation in leaf physiological strategies, from more conservative (low SLA) to more productive (high SLA) (Wright \emph{et al.} 2004). In the context of drought stress, low SLA might be adaptive as lower leaf surface area is expected to reduce water loss through transpiration. In several cases, reductions in SLA have been reported under drought conditions (Casper \emph{et al.} 2001). However, for all of these traits, the shape of plastic responses to variation in water availability remain largely unmeasured.

Less is known about how trait variation and relationships between traits change across continuous environmental gradients (but see Robinson \emph{et al.} 2009). Genetic variation for traits can be higher, for example, in less frequently encountered environmental conditions (Gibson and Dworkin 2004; Schlichting 2008; Paaby and Rockman 2014). And describing this structure is important as trait variances and covariances influence evolutionary constraints. For example, if total genetic variance in trait space changes depending on the environment, then the capacity to respond to selection will vary accordingly, with reduced responses to selection under conditions where genetic variance is lower and vice versa. Similarly, if trait covariances depend on environment (Wood and Brodie III 2015), conditions which increase trait covariances may limit evolutionary potential across a range of environments by reducing the effective axes of variation (Levins 1968; Lande and Arnold 1983; Via and Lande 1985; Kingsolver and Gomulkiewicz 2003; Gomulkiewicz \emph{et al.} 2018). These phenomena are made more complex by the possibility that the relationship between trait variances and covariances with the environment may be non-linear. Investigating the genetic architecture of multiple traits is therefore useful for understanding evolution in rapidly changing environments. Fortunately, this need has been met with the development of statistical methods to reduce the dimensionality of genetic variance covariance matrices and produce meaningful summaries describing evolutionary constraints of multiple phenotypes simultaneously (Houle 1992; Blows 2007; Kirkpatrick 2009; Kingsolver \emph{et al.} 2015).

To study the shape of phenotypic responses to the environment, modeling traits as mathematical functions of continuous variables, or function-valued-traits, has proven powerful (Kirkpatrick and Heckman 1989; Kingsolver \emph{et al.} 2001; Griswold \emph{et al.} 2008; Stinchcombe \emph{et al.} 2012; Goolsby 2015; Gomulkiewicz \emph{et al.} 2018). Recent decades have seen this approach used across many organisms (Gibert \emph{et al.} 1998; Pettay \emph{et al.} 2008; Rocha and Klaczko 2012; Mason \emph{et al.} 2020), traits (Robinson \emph{et al.} 2009; McGuigan \emph{et al.} 2010; Stinchcombe \emph{et al.} 2010), and components of the environment (Brommer \emph{et al.} 2008; McGuigan 2009; Pearse \emph{et al.} 2019). For plant scientists, experimental and analytical approaches leveraging the concept of function-valued-traits provide a framework for gaining a deeper understanding of plant acclimation and evolutionary adaptation to the environment.

In this study we combined these approaches into an exploratory investigation of trait plasticity, covariance and evolutionary constraints across soil moisture gradients. We model trait responses to a continuous gradient of soil moisture for multiple genotypes of two species of the model grass genus \emph{Brachypodium}: annual \emph{B. distachyon} and perennial \emph{B. sylvaticum}. Because the shapes of trait responses cannot necessarily be known \emph{a priori}, we use model selection among linear and non-linear environmental predictors to estimate the response function for each trait. We then estimate genotype means at different levels of soil moisture and compute the variance-covariance parameters for all traits. Finally, we ask whether patterns of variance and covariance in drying-responsive traits in \emph{Brachypodium} species may lead to variation in evolutionary constraint as a function of soil moisture.

@@

\emph{Brachypodium} is a model genus for the genetics and genomics of C3 grasses (Brkljacic \emph{et al.} 2011). In this work we studied natural variation between and among two species of \emph{Brachypodium}: the annual \emph{B. distachyon} and the perennial \emph{B. sylvaticum}. Both species are endemic to Eurasia, with \emph{B. distachyon} more prevalent in seasonally dry habitats in Southern Europe, North Africa, and the Middle East, and \emph{B. sylvaticum} more widely distributed throughout Eurasia (Catalan \emph{et al.} 2016) (Figure \ref{fig:maps}). The different evolutionary histories, ecological environments, and life habits of these two species may have impacted patterns of genetic variation and evolvalbility in relation to soil moisture.

To investigate genetic variation, five genotypes of each were studied to characterize patterns of variation in plant traits across an environmental gradient: \emph{Brachypodium distachyon} inbred lines ABR2, Adi-10, Bd21, Bd3-1, and Koz-1 and \emph{Brachypodium sylvaticum} inbred lines Ain-1, Ast-1, Kry-1, Osl-1, Vel-1. For each species, these genotypes represent a range of geographical origins and phenotypic diversity (Personal observations) (Steinwand \emph{et al.} 2013; Des Marais \emph{et al.} 2017). For example, Kry-1 is native to Krym, Ukraine while Ain-1 is native to Ain Draham, Tunisia (Steinwand \emph{et al.} 2013). Both species are self-compatible and each of the lines used here have been maintained as inbred lines for greater than six generations (Vogel \emph{et al.} 2009; Steinwand \emph{et al.} 2013); as such, experimental replicates may be considered nearly isogenic.

@@

Dry down treatments began 29 and 42 days after sowing (DAS) for \emph{B. distachyon} and \emph{B. sylvaticum}, respectively. Because the harvesting was divided over five consecutive days (see section below), plants were split into five equal harvest cohorts, with each cohort containing equal numbers of each watering treatment to avoid confounding harvest day with soil moisture content. Thus, though each consecutive cohort differed in age by a single day, each experienced the dry down treatment for the same amount of time. Nevertheless, we expected the difference in age between harvest cohorts to potentially impact trait expression and we therefore included harvest day (cohort) as a covariate in subsequent models. To generate a continuous gradient of final soil moisture by the end of the dry down period, plants were split into five watering treatments, receiving approximately 0, 4, 8, 12, 16, or 20 ml of water per day for 14 days (Figure \ref{fig:racks}). Prior to initiating the experiment pots were weighed with dry soil (\(mass_{dry}\)) and field capacity (\(mass_{max}\)). During the course of the dry down, soil moisture content was calculated during the morning of day \(d\) for each pot as \(mass_d/(mass_{max}-mass_{dry})\). The final soil moisture reflects the combined effects of water input and output through evaporation from the soil and evapotranspiration from the plants.

@@

To characterize phenotypic responses to our experimental soil moisture gradient, we measured a suite of developmental and physiological traits. Plants were harvested in five cohorts over five days at the end of the dry down period. Each day, half of the sampled plants were harvested for above and below ground biomass, total above ground green area, \(\delta C_{13}\), N content, C content. The other half were harvested and assessed for specific leaf area (SLA) and relative water content (RWC). Above ground leaf area was estimated by laying freshly harvested plants flat between plates of clear acrylic and imaging with a Nikon 5300 digital camera at a fixed distance with a 35mm Nikkor lens. Total green pixels were counted for each image with Easy Leaf Area (Easlon and Bloom 2014) (\url{https://github.com/heaslon/Easy-Leaf-Area}) with settings shown in Figure \ref{fig:leafarea}. Above ground biomass was measured after drying leaf material overnight at 60°C and then for several weeks at room temperature. Below ground biomass was measured after washing the soil matrix from roots and drying them overnight at 60°C and then for several weeks at room temperature. Above and below ground biomass was measured after leaves and roots were dried. Leaf tissues for \(\delta C_{13}\), \(\delta N_{15}\), nitrogen (hereafter \enquote{N}) content, and carbon (hereafter \enquote{C}) content were ground to a fine powder and processed by the UC Davis Stable Isotope Facility. SLA was calculated by scanning the two youngest fully emerged leaves with a 1 \(cm^2\) red square. Leaf area in \(mm^2\) was calculated from these same images using Easy Leaf Area. SLA was calculated as \(leaf\ area / biomass_{dry}\). These leaves were also used to calculate RWC. Prior to drying, fresh leaves were weighed (\(biomass_{fresh}\)) and then submerged under water in 15mL falcon tubes for several hours. They were then weighed (\(biomass_{turgid})\), oven-dried overnight, and weighed again (\(biomass_{dry}\)). RWC was calculated as \((biomass_{fresh}-biomass_{dry})/(biomass_{turgid}-biomass_{dry})\).

\begin{table*}[tbp]
\begin{center}
\begin{threeparttable}
\caption{\label{tab:models}Model selection. 'x' indicates parameters chosen in best minimum AIC model. H = harvest day, G = genotype, E = final soil moisture, E\textasciicircum{}2 = quadratic parameter, s(E) = 2nd degree natural spline parameter}
\scriptsize{
\begin{tabular}{ccccccccc}
\toprule
 & \multicolumn{1}{c}{G} & \multicolumn{1}{c}{E} & \multicolumn{1}{c}{E\textasciicircum{}2} & \multicolumn{1}{c}{s(E)} & \multicolumn{1}{c}{H} & \multicolumn{1}{c}{G*E} & \multicolumn{1}{c}{G*E\textasciicircum{}2} & \multicolumn{1}{c}{G*s(E)}\\
\midrule
RWC - B. distachyon &  &  & x & x & x &  &  & \\ \midrule
RWC - B. sylvaticum & x &  & x & x &  &  &  & x\\ \midrule
SLA - B. distachyon & x &  & x & x & x &  & x & \\ \midrule
SLA - B. sylvaticum & x &  &  & x &  &  &  & \\ \midrule
Green Area - B. distachyon & x &  &  & x & x &  &  & \\ \midrule
Green Area - B. sylvaticum & x &  & x &  & x &  & x & \\ \midrule
Shoot Mass - B. distachyon & x &  & x &  & x &  &  & \\ \midrule
Shoot Mass - B. sylvaticum & x &  & x &  & x &  &  & \\ \midrule
Root Mass - B. distachyon & x &  & x & x & x &  &  & x\\ \midrule
Root Mass - B. sylvaticum & x &  &  &  &  &  &  & \\ \midrule
Root:Shoot - B. distachyon & x &  &  & x & x &  &  & \\ \midrule
Root:Shoot - B. sylvaticum & x &  & x &  & x &  & x & \\ \midrule
Biomass - B. distachyon & x &  &  & x & x &  &  & \\ \midrule
Biomass - B. sylvaticum & x &  &  &  & x &  &  & \\ \midrule
C content - B. distachyon &  &  & x &  &  &  &  & \\ \midrule
C content - B. sylvaticum & x &  &  & x &  &  &  & \\ \midrule
d13c - B. distachyon & x &  & x &  & x &  &  & \\ \midrule
d13c - B. sylvaticum & x &  &  & x & x &  &  & \\ \midrule
N content - B. distachyon & x &  & x &  & x &  &  & \\ \midrule
N content - B. sylvaticum & x &  &  & x &  &  &  & x\\ \midrule
d15n - B. distachyon & x &  &  & x & x &  &  & \\ \midrule
d15n - B. sylvaticum &  &  & x & x & x &  &  & \\ \midrule
C:N ratio - B. distachyon & x &  & x &  & x &  &  & \\ \midrule
C:N ratio - B. sylvaticum & x &  & x &  &  &  & x & \\ \midrule
\bottomrule
\end{tabular}
}
\end{threeparttable}
\end{center}
\end{table*}

@@

\[v_T = \sum_{i=1}^{n} \lambda_i\]
